WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-06 articles/2014/Article-WSClean/wsclean.tex
New results and article almost finished.

\subsection{Discretization of w} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Original sources are 1~Jy, ghost sources are approximately 20\%. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article? -- could be explained by having more samples at smaller $w$s. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. [TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be of interest to do full-sky imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g. when searching for transients), or be useful for self-calibration or full-sky deconvolution. For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and selected this as the default for WSClean.

@@

TODO: for fairness, also check making 4 polarization images.

\subsection{Derivation of computing cost formulae}
Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the CASA and WSClean measurements using least-squares fitting. Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors. The measurements are weighted with the inverse standard deviation. The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.

For WSClean's time cost $t_\textrm{WSClean}$ we find 
\begin{align}\notag
t_\textrm{WSClean}& (w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag
& 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
\label{eq:wsclean-computing-cost}%
& 0.252 N_\textrm{vis} - 7.64
\end{align}
and for CASA we find
\begin{equation}
t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
\label{eq:casa-computing-cost}%
\end{equation}
Parameter $w_e$ is the maximum $w$ value and is estimated with
\begin{equation}
 w_e = \frac{\max |\mathbf{x}| \sin za + \max z \cos za}{\lambda \max |\mathbf{x}|} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right),
\end{equation}
where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

TODO: actually, $w_e$ should not be normalized by baseline length $\max z$: fix in fits too.

TODO: derive function for $t_\textrm{CASA} = t_\textrm{WSClean}$.

It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more general the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w$.

@@

TODO: calculate when to use w-snapshots
TODO: calculate optimum splitting time?

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are often referred to as $X$ and $Y$ because many software treats the two polarizations as such. However, this is not correct because $X$ is officially defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output real valued $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

\section{Discussion} \label{sec:discussion}
WSClean has been build out of necessacity to be able to make images with the MWA, because no

TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)

TODO: discuss at what is the optimal time to split measurement sets into when using recentering technique.

TODO: A-projection makes the psf more invariant: therefore, truely ``joined'' deconvolution might be more accurate.

TODO: discuss low w-values means less-direction-invariant PSF.

>>>>

\subsection{Discretization of w} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article? -- could be explained by having more samples at smaller $w$s. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. [TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because they are scientifically of interest (e.g. when searching for transients). Full-sky imaging can also be useful for self-calibration or deconvolution. For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.

@@

It is important to point out that all these measurement were performed by imaging a single polarization. Making full-polarization images with WSClean currently requires four runs. Imaging four Stokes parameters with CASA costs slightly less than four times the time of imaging a single polarization. With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.

\subsection{Derivation of computing cost formulae}
Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the CASA and WSClean measurements using least-squares fitting. Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors. Measurements are weighted with the inverse standard deviation. The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.

For the WSClean time cost function $t_\textrm{WSClean}$ we find 
\begin{align}\label{eq:wsclean-computing-cost}
& t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
& N_\textrm{chan} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
\end{align}
and for CASA we find
\begin{align}\label{eq:casa-computing-cost}
& t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
& N_\textrm{chan} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
\label{eq:casa-computing-cost}%
\end{align}
Parameter $w_e$ is the maximum $w$ value and is estimated with
\begin{equation}
 w_e = \frac{1}{\lambda}\left[\max |\mathbf{x}| \sin \textrm{ZA} + \max z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
\end{equation}
where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w$.

TODO fill in formulae for common surveys, e.g. ESKAP and LOFAR.

@@

TODO: calculate when to use w-snapshots: discuss at what is the optimal time to split measurement sets into when using recentering technique.
TODO: calculate optimum splitting time?

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

\section{Discussion \& conclusions} \label{sec:discussion}
We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations. The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, and already an order of magnitude faster at a relatively small zenith angle of $10$\degree. At the same time, it is slightly more accurate. For approximately zenith angles $>15\degree$ or field of views $>35\degree$, rephasing visibility sets and performing snapshot imaging becomes worthwhile. This can lead to performance improvements of a factor of $3$ for the MWA. For arrays with longer baselines and higher resolution, snapshot imaging leads to even higher performance improvements. This agrees with \citet{widefield-imaging-ska-cornwell}. Additionally, we have shown that to implement $w$-snapshot imaging, a pure $w$-stacking implementation is likely more efficient than pure $w$-projection. A hybrid between the two might improve things by a few factors. TODO should we calculate some timings with SKA's configuration?

It remains challenging to joinedly deconvolve observations that are long enough to encur significant changes in the beam shape. It is possible to clean such an observation without beam corrections, but then a constant amount of flux is subtracted over time, which leads to inaccuracies. Alternatively, beam corrections can be made, either by integrating beam-corrected snapshots or by performing $a$-projection. This however leads to a stronger spatially-varying point-spread function, and would allow fewer image-based minor iterations in a major cycle. When it is required to calculate the beam on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes much more expensive. For a few hundred components, an approach with direct Fourier transforms is accurate and affordable, but this becomes intractable for thousands of components. Images with diffuse emmission typically require cleaning of more than ten-thousand components. Joined deconvolution is not an issue for achieving low to intermediate dynamic range with the MWA. Because of the excellent instanteneous $uv$-coverage and near-confusion noise level, snapshots can be cleaned individually to deep levels. For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations. However, MWA's EoR observations (TODO cite) require more advanced deconvolution techniques.
